Polynomial Time and Full Dependent Types

### 2023-07-05 paper/paper.tex
[ paper ] Some work on describing the realisability model

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Implicit Polynomial Time and Dependent Types}

@@

The story:
\begin{enumerate}
\item There are systems for implicit computational complexity based on
  linear logic (e.g. LFPL, SAL, LLL, BLL) that seek to capture
  complexity classes (most often polynomial time) implicitly
  (i.e. without directly talking about time bounds). This work has
  also led to followup work on explicitly resourced calculi such as
  RAML, which seek to use static analysis to determine complexity
  bounds on programs.
\item In dependent type theory, we can reason about the extensional
  behaviour of programs (within the limits of intensional type
  theory), but not about intenstional properties. The only way to
  formalise something like ``for all polynomial time functions
  $\mathbb{N} \to \mathbb{N}$'' is to develop a deep emebdding of
  polynomial time functions and to work with that, which is tedious.
\item The goal of this paper is to unify dependent types for reasoning
  about programs with implicit computational complexity type
  systems. In the first instance, we do not seek to use dependent
  types to state precise bounds on programs' time consumption, but we
  intend to use dependent types to reason about the extensional
  behaviour of programs that we know to be polynomial time. At the end
  of the paper, we show how to adapt our framework to explicit
  resource accounting, in the style of RAML.
\item The plan of the paper is as follows:
  \begin{enumerate}
  \item We first introduction the ideas behind the use of linear logic
    for capturing polynomial time, by studying two systems Lafont's
    SAL, and Hofmann's LFPL. In Hofmann and Dal Lago's presentation,
    these are two variants of Second order Multiplicative Linear
    Logic, with restricted $\oc$ modalities for duplication. We also
    recall the realisability semantics of Dal Lago and Hofmann (FIXME:
    move this later?)
  \item The most modern presentations of these systems use second
    order quantification to represent data types. This is incompatible
    with the form of dependent types we use here, so we seek a direct
    representation of natural numbers that captures
  \item We then recall QTT, a extension of Martin-Löf type theory with
    linear types, which can accomodate intensional restrictions on
    computability as well as full reasoning about programs. Since
    allowing an unrestricted $\oc$ modality would violate our
    polynomial time bounds, we instead choose the natural number
    semiring for resource accounting.
  \item We then transplant the simply typed rules into QTT, showing
    how to carefully distinguish between data that is available to the
    program, and data that is only available for type formation.
  \item We finish by showing that we can also do explicit resource
    tracking.
  \end{enumerate}
\end{enumerate}

@@

@@

\begin{definition}
  A \emph{resource monoid} $M$ consists of:
  \begin{enumerate}
  \item A carrier set $|M|$, whose elements represent amounts of
    potential. We use Greek letters $\alpha$, $\beta$, $\gamma$ to
    denote elements of a resource monoid.
  \item Commutative monoid structure $(\oplus,\emptyset)$ on $|M|$, so
    we can add potentials.
  \item A \emph{difference function} $M : |M| \times |M| \to \natinf$,
    where $\natinf$ is the natural numbers extended with a negative
    infinity $- \infty$ and $- \infty + k = -\infty$. A difference
    $M(\alpha, \beta) = k \in \mathbb{N}$ means that starting with
    potential $\alpha$ and ending with potential $\beta$ yields $k$
    units of fuel. A difference of $- \infty$ means that $\alpha$
    contains insufficient potential to reach $\beta$. Differencing
    must satisfy:
    \begin{enumerate}
    \item for all $\alpha$, $M(\alpha,\alpha) = 0$; and
    \item for all $\alpha, \beta, \gamma$,
      $M(\alpha, \beta) + M(\beta, \gamma) \leq M(\alpha, \gamma)$.
    \end{enumerate}
    The latter is a ``reverse triangle inequality'': the fuel
    recoverable by moving between potential levels $\alpha$ and
    $\gamma$ via $\beta$ may be less than the fuel recoverable
    moving from $\alpha$ to $\gamma$ directly.
  \item An \emph{accounting function}
    $\mathit{acct} : \mathbb{N} \to |M|$ such that for all $k$,
    $k \leq M(\mathit{acct}(k),\emptyset)$.
  \end{enumerate}
\end{definition}

@@

@@

\subsection{Realisability Model}
\label{sec:realisability-model}

\bob{Rearrange this as an indexed linear poset}

Equipped with our underlying costed model of computation
(\autoref{sec:machine-model}) and a compositional notion of resource
potential (\autoref{sec:resource-monoids}), we now begin to construct
models of QTT that witness the resource (and type) soundness of our
complexity constrained systems.

\bigskip

\cite{atkey18qtt} describes a general class of QTT models termed
\emph{Quantitative Categories with Families} (QCwFs), an extension of
the \cite{dyber}'s \emph{Categories with Families} formulation of
models of dependent type theory. We will not regurgitate the full
definition here. Instead, we will define each of the parts of our
specialised QCwF for proving complexity soundness and explain
informally how to use them to model the syntax of QTT. Please refer to
\cite{atkey18qtt} for details of the interpretation.

Fix a resource monoid $M$ with a submonoid $M_0 \subseteq M$.

We define a category $\mathbf{Asm}(M,M_0)$.\bob{``Asm'' is the wrong
  name here.}

\begin{definition}[Resourced Sets, Semantic Contexts and Types]
  \label{defn:rl-object}
  \begin{enumerate}
  \item A \emph{resourced set} $X$ is a pair $(|X|,\models_\Gamma)$ of
    a set $|X|$ and a ternary relation
    $\mathord{\models_\Gamma} \subseteq |M| \times \ValSet \times |X|$. We write
    the first two elements to the left of the $\models$ and the last
    to the right: e.g. $\alpha, v \models_X x$.
  \item \emph{Semantic contexts} $\Gamma$ are resourced sets.
  \item Given a set $\Delta$ (representing a context bereft of any
    resource information), a family $S$ of semantics types over
    $\Delta$ is a collection $\{S_\delta\}_{\delta \in \Delta}$ of
    resourced sets. We write $\Ty(\Delta)$ for the large set of all
    semantic types over $\Delta$.
  \end{enumerate}
\end{definition}
Note how the intepretation of types only depends on a set of values,
not any resource information. This corresponds to

We define scaling and addition of resourced sets (and also semantic
contexts) to interpret the scaling and addition of syntactic contexts
\autoref{defn:FIXME}.

\begin{definition}[Scaling and Addition of Resourced Sets]
  \begin{enumerate}
  \item \emph{(Scaling)} If $X$ is a resourced set and
    $n \in \mathbb{N}$, then $nX$ is the resourced set with
    $|nX| = |X|$ and $\alpha, v \models_{nX} x$ iff there exist
    $\alpha_1, \dots, \alpha_n$ such that
    $\alpha \leq \alpha_1 \oplus \dots \oplus \alpha_n$ and for all
    $1 \leq i \leq n$, $\alpha_i, v \models_X x$.
  \item \emph{(Addition)} If $X$ and $Y$ are resourced sets such that
    $|X| = |Y|$, then define their addition $X + Y$ with
    $|X + Y| = |X|$ and $\alpha, v \models_{X + Y} x$ iff there exist
    $\alpha_1, \alpha_2$ such that
    $\alpha \leq \alpha_1 \oplus \alpha_2$ and
    $\alpha_1, v \models_X x$ and $\alpha_2, v \models_Y x$.
  \end{enumerate}
\end{definition}

Morphisms are where the connection between abstract elements, machine
values, and potential is given force:
\begin{definition}[Morphisms, Semantic Substitutions and Terms]
  \label{defn:rl-morphism}
  \begin{enumerate}
  \item Given resourced sets $X$ and $Y$, a morphism $f : X \to Y$ is
    a function $f : |X| \to |Y|$ for which there exists a closed
    expression $e \in \ExpSet$ (the \emph{realiser}) and a potential
    $\gamma \in |M_0|$ such that for all $\alpha, v, x$ with
    $\alpha, v \models_X x$, there exists a $\beta$, $v'$ and $k$ such
    that:
    \begin{enumerate}
    \item $e, [v] \Downarrow_k v'$ (\emph{the realiser evaluates to a result value}) and
    \item $k \leq M(\alpha + \gamma, \beta)$ (\emph{the potential difference yields adequate fuel}) and
    \item $\beta, v' \models_Y f(x)$ (\emph{the result is correct}).
    \end{enumerate}
  \item A \emph{semantic substitution} $\Gamma \Rightarrow \Delta$ is
    a morphism between the resourced sets.
  \item Given a semantic context $\Gamma$ and a semantic type
    $S \in \Ty(|\Gamma|)$, a \emph{semantic resourced term} $m$ is a
    function $m : (\gamma \in |\Gamma|) \to |S(\gamma)|$ such that there
    exists a closed expression $e \in \ExpSet$ and potential
    $\gamma \in |M_0|$ satisfying a similar property to that for
    morphisms. We write $\RTm(\Gamma, S)$ for the set of all semantic
    resourced terms between $\Gamma$ and $S$.
  \item Given a set $\Delta$ and a semantic type $S \in \Ty(\Delta)$,
    a \emph{semantic unresourced term} $m$ is a function
    $m : (\delta \in \Delta) \to |S(\delta)|$. We write
    $\Tm(\Gamma, S)$ for the set of all semantic unresourced terms
    between $\Delta$ and $S$.
  \end{enumerate}
\end{definition}
The definition of \emph{semantic resourced term} is the key to our
claim of soundness for any applications of QTT. By choosing the
correct resource monoid, and interpreting QTT with extensions
supported by that monoid into the resulting structure, we can deduce
our desired resource soundness theorems. FIXME: forward references.

Notice that the resource potential $\gamma$ for morphisms and
resourced terms must come from the sub-monoid $M_0$. For the
polynomial time systems, this will ensure that programs cannot
generate iterability from scratch. This is benign in the case of the
cons-free system (though it does complicate the statement of
soundness), but would be catastrophic in the LFPL-style system because
it would mean that we could generate $\Diamond$s from nothing.

To be able to interpret

\begin{definition}
  Context extension
\end{definition}

\subsection{Realising Finite Type Formers}

\begin{definition}[$\Pi$-types]
  FIXME: taken from
\end{definition}

\begin{definition}[$\Sigma$-types]

\end{definition}

\begin{definition}[Equality types]

\end{definition}

\begin{definition}[Non-iterable types]

\end{definition}

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Polynomial Time and Full Dependent Types}

@@

{\bf New story:}
\begin{enumerate}
\item Introduction: The goal of this paper is to unify implicit
  computational complexity with full dependent types. ICC enables us
  to ensure that \emph{programs} are within a certain complexity
  class, while dependent types allow us to state and prove that these
  programs solve the problems that we intend them to, and to construct
  reductions between problems. We can marry complexity-controlled
  implementations with problems, paving the way for a \emph{synthetic
    computational complexity theory}.
\item Section 1: Linear types and Implicit Polynomial Time.
  \begin{enumerate}
  \item Total typed functional programming, as we see in type theory,
    is restricted to the use of higher order functions and iteration
    over inductive data types.
  \item Polynomial time is the smallest class closed under nested
    iteration over the input -- i.e. we consider iteration over the
    input as feasible, and compositions thereof. The key thing is to
    avoid iteration over data constructed from the input.
  \item Key is to control iteration. So we split inductive datatypes
    into iterable and non-iterable versions. We can
  \item One way to capture polytime is to ensure that all iteration is
    only done over the input. So we can duplicate iterable things, but
    we can't construct new iterable things (at least not from an
    iteration). We also can't duplicate higher order things, because
    that effectively acts as a way to duplicate iterations (half
    instantiated iterators). Compare to Jones' CONS-FREE programming.
  \item Another approach is to allow ``borrowing'' of iterability by
    means of diamonds. Iterability becomes a first-class object, so we
    can use it to construct new data structures from the input. This
    allows the construction of new iterable data structures. LFPL.
  \end{enumerate}
\item Section 2: Extending to Dependent Types
  \begin{enumerate}
  \item Combining linear types and dependent types is subtle, because
    there is a fundamental incompatibility in the syntax of the two
    systems.
  \item There have been several approaches, though they are all
    somewhat similar (list them all here). We will use QTT, because it
    offers a firm semantic foundation as well as a relatively
    straightforward syntax.
  \item Recap the syntax of QTT, with extensional equality. Key point
    is that there is a normal dependently typed language with a linear
    sub-language. We can have special constructs in the linear
    sub-language that ``erase'' to normal type theory -- products are
    the prime example.
  \item Non iterable datatypes -- have normal eliminators in the DTT
    fragment, but
  \end{enumerate}
\item Section 3: Programming and Proving with Poly-time
  \begin{enumerate}
  \item Internalised completeness proofs.
  \item Construct the type of problems that are solvable in polynomial
    time.
  \item Construct reductions between problems.
  \item NP. Doesn't make sense to talk about this without dependent
    types, because we can't say when we've solved the problem being
    stated.
  \item PP (Probabilistic), using a monad. Again, need to be able to
    state connection with the problem to know that we are solving it.
  \end{enumerate}
\item Section 4: Soundness via Realisability
  \begin{enumerate}
  \item The underlying CBV language and its
  \end{enumerate}
\item Section 5: Extensions
  \begin{enumerate}
  \item Explicit Complexity via Diamond Counting
  \item Poly-time univalence
  \item Towards P- and NP-completeness proofs. Issue here is the
    extensionality. Towards a synthetic (poly-time based)
    computational complexity theory.
  \end{enumerate}
\item Section 6: Related Work
  \begin{enumerate}
  \item ICC stuff
    \begin{enumerate}
    \item LFPL
    \item CONS-free
    \item Soft Affine Logic; which has the same underlying idea as
      CONS-free (is this a novel observation?)
    \item Other systems based on information flow.
    \end{enumerate}
  \item Explicit complexity: lambda-amor, and things by Deepak Garg
    and Marco Gaboardi.
  \item Recent CMU stuff on calf
  \end{enumerate}

\end{enumerate}

{\bf Old story:}
\begin{enumerate}
\item There are systems for implicit computational complexity based on
  linear logic (e.g. LFPL, SAL, LLL, BLL) that seek to capture
  complexity classes (most often polynomial time) implicitly
  (i.e. without directly talking about time bounds). This work has
  also led to followup work on explicitly resourced calculi such as
  RAML, which seek to use static analysis to determine complexity
  bounds on programs.
\item In dependent type theory, we can reason about the extensional
  behaviour of programs (within the limits of intensional type
  theory), but not about intenstional properties. The only way to
  formalise something like ``for all polynomial time functions
  $\mathbb{N} \to \mathbb{N}$'' is to develop a deep emebdding of
  polynomial time functions and to work with that, which is tedious.
\item The goal of this paper is to unify dependent types for reasoning
  about programs with implicit computational complexity type
  systems. In the first instance, we do not seek to use dependent
  types to state precise bounds on programs' time consumption, but we
  intend to use dependent types to reason about the extensional
  behaviour of programs that we know to be polynomial time. At the end
  of the paper, we show how to adapt our framework to explicit
  resource accounting, in the style of RAML.
\item The plan of the paper is as follows:
  \begin{enumerate}
  \item We first introduction the ideas behind the use of linear logic
    for capturing polynomial time, by studying two systems Lafont's
    SAL, and Hofmann's LFPL. In Hofmann and Dal Lago's presentation,
    these are two variants of Second order Multiplicative Linear
    Logic, with restricted $\oc$ modalities for duplication. We also
    recall the realisability semantics of Dal Lago and Hofmann (FIXME:
    move this later?)
  \item The most modern presentations of these systems use second
    order quantification to represent data types. This is incompatible
    with the form of dependent types we use here, so we seek a direct
    representation of natural numbers that captures
  \item We then recall QTT, a extension of Martin-Löf type theory with
    linear types, which can accomodate intensional restrictions on
    computability as well as full reasoning about programs. Since
    allowing an unrestricted $\oc$ modality would violate our
    polynomial time bounds, we instead choose the natural number
    semiring for resource accounting.
  \item We then transplant the simply typed rules into QTT, showing
    how to carefully distinguish between data that is available to the
    program, and data that is only available for type formation.
  \item We finish by showing that we can also do explicit resource
    tracking.
  \end{enumerate}
\end{enumerate}

@@

\newcommand{\rplus}{\oplus}
\newcommand{\rzero}{\emptyset}

FIXME: introductory paragraph.

@@

\begin{definition}
  A \emph{resource monoid} $M$ consists of:
  \begin{enumerate}
  \item A carrier set $|M|$, whose elements represent amounts of
    potential. We use Greek letters $\alpha$, $\beta$, $\gamma$ to
    denote elements of a resource monoid.
  \item Commutative monoid structure $(\rplus,\rzero)$ on $|M|$, so
    we can add potentials.
  \item A \emph{difference function} $M : |M| \times |M| \to \natinf$,
    where $\natinf$ is the natural numbers extended with a negative
    infinity $- \infty$ and $- \infty + k = -\infty$. A difference
    $M(\alpha, \beta) = k \in \mathbb{N}$ means that starting with
    potential $\alpha$ and ending with potential $\beta$ yields $k$
    units of fuel. A difference of $- \infty$ means that $\alpha$
    contains insufficient potential to reach $\beta$. Differencing
    must satisfy:
    \begin{enumerate}
    \item for all $\alpha$, $M(\alpha,\alpha) = 0$; and
    \item for all $\alpha, \beta, \gamma$,
      $M(\alpha, \beta) \rplus M(\beta, \gamma) \leq M(\alpha, \gamma)$.
    \end{enumerate}
    The latter is a ``reverse triangle inequality'': the fuel
    recoverable by moving between potential levels $\alpha$ and
    $\gamma$ via $\beta$ may be less than the fuel recoverable
    moving from $\alpha$ to $\gamma$ directly.
  \item \bob{Interaction between $(\rplus,\rzero)$ and $M(\alpha,\beta)$.}
  \item An \emph{accounting function}
    $\mathit{acct} : \mathbb{N} \to |M|$ such that for all $k$,
    $k \leq M(\mathit{acct}(k),\rzero)$.
  \end{enumerate}
\end{definition}

@@

\bob{PLAN: RCwFs, and state that they can be constructed from Indexed Linear Posets}

@@

\bob{define the construction}
\begin{enumerate}
\item Start with a CwF $\cat{C}$ that supports $\Pi$, $\Sigma$ and
  natural numbers.
\item Assume a functor $L : \cat{C}^\top \to \LinPoset$ with
  reindexing along projections having a right adjoint.
\item Construct a new CwF with the same base category as $\cat{C}$,
  but with $\Ty'(\Delta) = \Sigma A : \Ty(\Delta).~L(\Delta.A)$ and
  $\Tm'(\Delta, (T, X)) = \Tm(\Delta, T)$.
\item
\end{enumerate}

This construction is implicit in \cite{atkey2018qtt}, where the
indexed linear poset is defined from a $\mathcal{R}$-LCA.

\subsection{Amortised Complexity Realisability Model}
\label{sec:realisability-model}

Equipped with our underlying costed model of computation
(\autoref{sec:machine-model}) and a compositional notion of resource
potential (\autoref{sec:resource-monoids}), we construct models of QTT
that witness the resource and type soundness of our complexity
constrained systems.

\bob{Tie back to previous section}

We fix a resource monoid $M$ with distinguished sub-monoid $M_0$.

\subsubsection{Base CwF}
\label{sec:realisability-base-cwf}

As our base CwF, we use the category $\Set$ of sets and functions with
the standard interpretation of $\Pi$, $\Sigma$ types and the
$\mathbb{N}$ type.

\subsubsection{Indexed Linear Preorder}
\label{sec:realisability-indexed-linear-preorder}

We now define an indexed linear poset $L$ of realisers over $\Set$
that ties together our ``mathematical'' model of types in $\Set$ with
our machine model and resource monoid. For a set $A$, the carrier of
$L(A)$ is the set of ternary relations
$X \subseteq A \times M \times \ValSet$ and we define the ordering
$X \leq Y$ to hold iff there exists realising expression
$E \in \ExpSet$ and potential $\gamma \in M_0$ such that for all
$a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
$(a,\alpha,v) \in X$, we have that there exists a result
$v' \in \ValSet$, step count $k \in \mathbb{N}$ and result potential
$\beta \in M$ with:
\begin{enumerate}
\item $E, v \Downarrow_k v'$ (evaluation successfully completes in $k$ steps);
\item $(a, \beta, v') \in Y$ (the result is well-resourced and
  satisfies $Y$); and
\item $k \leq M(\alpha \rplus \gamma, \beta)$ (the step count is within the
  difference between the initial potential and the result potential).
\end{enumerate}
Note that the definition of realisablity is uniform in the element $a$
-- the realising expression $e$ and the potential $\gamma$ must work
for all $a$ -- thus the implementation and complexity measure of the
transition being modelled cannot depend on what the input is. Put in
implementation terms, the input $a$ is not present at
runtime. Moreover note that the potential $\gamma$ attached to the
expression $e$ must come from the sub-monoid $M_0$, indicating that is
intended to be data-free, while the potential $\alpha$ for the input
is from the full monoid $M$, so it can contain data and functions.

For $X, Y \in L(A)$, the required elements for symmetric monoidal
closed structure are defined as follows. For the tensor product
$X \otimes Y \in L(A)$, the realising value must be a pair $(v_1,v_2)$
and the potential of the pair must split into suitable potentials
$\alpha_1$, $\alpha_2$ for the components. For the residual
$X \multimap Y$, the realising value must be a closure with potential
to, when added to the potential of an input, compute the output with
enough remaining. Note that the potential attached to a closure
($\alpha$, here) need not be from the sub-monoid $M_0$. Unlike
top-level term interpretations, closures may contain data.
\begin{displaymath}
  \begin{array}{lcl}
    X \otimes Y &=& \{ (a, \alpha, (v_1, v_2)) \mid \exists \alpha_1, \alpha_2.~0 \leq M(\alpha, \alpha_1 \rplus \alpha_2) \land X(a,\alpha_1,v_1) \land Y(a,\alpha_2,v_2) \}\\
    X \multimap Y &=& \{ (a, \alpha, \clo{E}{\eta}) \mid
                      \begin{array}[t]{@{}l}
                        \forall \alpha' \in M, v,w \in \ValSet.\,X(a,\alpha',v) \Rightarrow\\
                        \quad \exists v', k, \beta.\,
                        E, (\eta :: w :: v) \Downarrow_k v' \land Y(a,\beta,v') \land k \leq M(\alpha \rplus \alpha', \beta) \} \end{array}
  \end{array}
\end{displaymath}
The seemingly useless $w \in \ValSet$ in the formula for
$X \multimap Y$ is a dummy argument standing for the self-referential
reference to the closure used for defining recursive programs.

Each $L(A)$ has a terminal (i.e. top) element, which is also the unit
for $\otimes$, defined as
$I_A = \{(a, \alpha, *) \mid a \in A, \alpha \in M\}$. The potential
$\alpha$ here is unrestricted, so $I_A$ can consume an arbitrary
resource.

$\mathbb{N}$-Graded exponentials in each $L(A)$ are defined using the
action $(\cdot)$ of $(\mathbb{N}, \leq)$ on $M$ defined above \bob{do
  this}. At the level of realising values, the modality $\oc_n$ has no
effect; it only serves to alter the required potentials.
\begin{displaymath}
  \oc_n\,X = \{(a,\alpha,v) \mid \exists \alpha'.\,M(n \cdot \alpha', \alpha) = 0 \land (a, \alpha', v) \in X \}
\end{displaymath}

$L$ also has arbitrary $\Set$-indexed products, realised ``lazily'' as
functions that take dummy arguments. For $A \in \Set$ and
$B \in A \to \Set$ and $X \in L(\Sigma A.\,B)$, we define
$\forall_B X \in L(A)$ similarly to $\multimap$ above, but with
different resource and indexing requirements:
\begin{displaymath}
  \forall_B X = \{ (a,\alpha,\clo{E}{\eta}) \mid \forall b, v.~\exists v', \beta, k. E, (\eta :: v :: *) \Downarrow_k v' \land X((a,b),\beta,v') \land k \leq M(\alpha,\beta) \}
\end{displaymath}
Note, as with the definition of $X \leq Y$ above, the realiser closure
$\clo{E}{\eta}$ must be chosen uniformly for all $b$. This definition
also appears to allow arbitrary computation (paid for by $\alpha$) to
happen when the realising closure is applied, but the potential
$\alpha$ will only ever be greater than $\beta$ by enough to handle
the administrative costs of applying the function.

To complete the construction of $L$ as an indexed linear preorder, we
need to give realisers for each of the required inequalities in
\autoref{def:indexed-linear-preorder}. In each case, this is a matter
of programming in the language of \autoref{sec:machine-model}. The
corresponding potentials are calculated by counting the steps in the
ensuing programs. Note that, so far, all realisers are constant time
(relative to their input), so the constructions so far work for any
resource monoid. For systems that require iteration, we will need more
structure on our resource monoids \bob{fwd ref}.

\begin{proposition}
  $L$, with $I$, $\otimes$, $\multimap$, $\oc_n$, and $\forall_{B}$ defined above, is
  an indexed linear preorder.
\end{proposition}

\begin{proof}
  See the Agda development.
\end{proof}

\begin{proposition}\label{prop:basic-qtt-realisability-model}
  Get a model QTT with $\Pi$, $\Sigma$, and a universe.
\end{proposition}

\subsubsection{Non-iterable Data Types}

The model of QTT constructed in
\autoref{prop:basic-qtt-realisability-model} does not yet include any
useful base types. Handling iterable types, which are the ones that
induce non-constant-time complexities, requires specific properties of
resource monoids that we introduce in
\autoref{sec:realising-iteration}.

Rather than perform a complex generic construction of datatypes to
show how non-iterable versions may be realised, we focus on two
representative examples. Booleans

\paragraph{Booleans} Fix $\mathbb{B} = \{ \mathit{tt}, \mathit{ff} \}$
as our set of boolean elements. We define an element of $L(\mathbb{B}$
to represent boolean values:
\begin{displaymath}
  \mathrm{bool} = \{ (\mathit{tt}, \alpha, \synTrue) \mid \alpha \in M \} \cup \{ (\mathit{ff}, \alpha, \synFalse) \mid \alpha \in M \}
\end{displaymath}
Thus, the boolean $\mathit{tt}$ is represented by the value $\synTrue$
and $\mathit{ff}$ is represented by $\synFalse$. In both case, we
allow arbitrary potentials to be attached.

\paragraph{Lists}
